### 2020-05-29 paper.tex
Answer RQ1

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs. E.g. when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question. Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.

@@

Salvaneschi et al. \cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of data. Think of a spreadsheet application\cite{10.1145/2501654.2501666} that links the values of cells to other cells using formulas. Once a cell is changed, it is the task of the application to update all dependant cells accordingly and cascade the updates throughout the model. This data-flow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.

@@

RP is a declarative programming paradigm. Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved. The languages runtime system then figures out a way to satisfy that given description. RP environments specifically provide abstractions to describe a data-flow and dependency graph which is used to transform and direct data at runtime.

@@

Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the data-flow graph. In fact, RP provides its own challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g. the predecessor transformation according to the data-flow graph. Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph. A \textbf{reactive debugger} has knowledge about the internal model of a RP runtime system. It leverages on it and provides specialized tools e.g. to navigate or visualize the data-flow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}. We will have a closer look to available reactive debugger solutions.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It has the ability to push (``emit'') values to its dependents. Multiple depending Observables form and describe a data-flow graph eventually. A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

All of the interviewees referred to RP with an overall positive experience. We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions. Hence a big strength of RP seems to be the ability to describe complex data-flow constructs using a specialized DSL. The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.

@@

The overall sentiment in both the informal interviews as well as the reported war stories in regards of RP with RxJS was positive. The software engineers value how they can describe data flows using a DSL, even though the learning curve was perceived as steep. We heard some interesting reports how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests. Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.

Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement data-flow graphs with minimal, formal correctness. We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though. Yet traditional debugging tools allowed them to inspect various parts of a stream, they were unable to get more information about e.g. life cycles of the Observable composition. Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem. Listing~\ref{lst:rxjs-debugging} exemplifies challenges when debugging a stream of Observables with traditional debugging tools.

@@

Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools. Most of them provide a way to represent the data flow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy). They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector). Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-related tools provide small, differentiated feature sets. Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).

@@

\section{Future Work}
\label{sec:future}

>>>>

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs. E.g. when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question. Assuming a dataflow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.

@@

Salvaneschi et al. \cite{7827078} define a RA as a computer program which processes a continuos, indefinite and ever-changing flow of data. Think of a spreadsheet application\cite{10.1145/2501654.2501666} that links the values of cells to other cells using formulas. Once a cell is changed, it is the task of the application to update all dependant cells accordingly and cascade the updates throughout the model. This dataflow based type of application is contrasted by program-flow based applications which consume and process one piece of input while producing a one-off output like an image format conversion program that transforms between HEIF\footnote{\url{https://nokiatech.github.io/heif/technical.html}} and JPEG image formats.

@@

RP is a declarative programming paradigm. Where engineers use imperative programing languages to tell a program every step \emph{how} to do something, a declarative language allows to describe \emph{what} wants to be achieved. The languages runtime system then figures out a way to satisfy that given description. RP environments specifically provide abstractions to describe a dataflow and dependency graph which is used to transform and direct data at runtime.

@@

Where individual transformations of an RP based program can be inspected using traditional debuggers, they fall short in understanding the underlying model of the dataflow graph. In fact, RP provides its own challenges to debuggers: Call stacks expose internal invocations of the RP runtime system rather than e.g. the predecessor transformation according to the dataflow graph. Further, breakpoints can only be used on the imperative parts of transformations and lack the functionality of interrupting execution when the RP runtime hits a specific node within the graph. A \textbf{reactive debugger} has knowledge about the internal model of a RP runtime system. It leverages on it and provides specialized tools e.g. to navigate or visualize the dataflow graph \cite{10.1145/2884781.2884815} \cite{10.1145/3180155.3180156} \cite{rxviz}. We will have a closer look to available reactive debugger solutions.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It has the ability to push (``emit'') values to its dependents. Multiple depending Observables form and describe a dataflow graph eventually. A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

All of the interviewees referred to RP with an overall positive experience. We heard statements like ``RP is a good way for composing multiple data sources'' and ``having a statically typed language combined with RP already guarantees some kind of basic correctness of my program'' repeatedly throughout all sessions. Hence a big strength of RP seems to be the ability to describe complex dataflow constructs using a specialized DSL. The learning curve can be steep for a novice engineer though: ``Being already challenged with new concepts of Angular and ngrx, I noticed RxJS concepts and operators to be hard to grasp in addition on top'' an interview partner giving lectures in frontend web application development stated further.

@@

The overall sentiment in both the informal interviews as well as the reported war stories in regards of RP with RxJS was positive. The software engineers value how they can describe dataflows using a DSL, even though the learning curve was perceived as steep. We heard some interesting reports how RxJS is applied in a daily development environment: Marble diagrams help to understand  an Observables behavior and are useful to implement tests. Large frameworks like Angular hide some of the complexity of RxJS but allow engineers to make use of its full power once the provided functionality needs to be extended.

Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement dataflow graphs with minimal, formal correctness. We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though. Yet traditional debugging tools allowed them to inspect various parts of a stream, they were unable to get more information about e.g. life cycles of the Observable composition. Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem. Listing~\ref{lst:rxjs-debugging} exemplifies challenges when debugging a stream of Observables with traditional debugging tools.

@@

Specialized utilities are aware of RP runtime environments and help software engineers to interpret its state better than before using sole traditional debugger tools. Most of them provide a way to represent the dataflow graph either graphically (Reactive Inspector, rxfiddle) or in textual form (rxjs-spy). They visualize the data flowing through the graph over time in marble diagrams (rxfiddle, Rx Visualizer) or directly in a visual graph representation (Reactive Inspector). Some like Reactive Inspector for REScala provide an ``all in one'' solution with breakpoints, omniscient debugging capabilities and full IDE integration whereas the RxJS-related tools provide small, differentiated feature sets. Most of them require the engineer to extract code from its origin (rxfiddle, Rx Visualizer) or require manual code preparation or modification (rxjs-spy).

@@

Talking to professionals during the informal reviews, evaluating the war stories as well as the hands-on experiences and results from the observational study allowed us to identify three main challenges software engineers have to overcome when working with RxJS within an application: (i) They want to understand how and what data flows through the graph described by the RP DSL. This helps to identify simpler problems most of the time, though knowing about the life cycle events of Observables was noted as crucial especially when working with more complex, higher-order compositions of Observables. (ii) Where traditional debugging tools provide support suited for an imperative programming style, they fall short when interacting with RP-based code which would require a deeper understanding of the RP runtimes internal graph-based model. This makes it hard for engineers to retrace dataflows and computations. And finally, (iii) even though there are partial solutions available to the two previous described challenges, they usually require the software engineers either to manually modify source code or extracting parts of it to an external utilities. These additional efforts cost time and bring their own risks since production code needs to be touched without adding actual value.

\section{Outlook}
\label{sec:future}
